The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2016-07-16 article.tex
Cleaned up the chem formula section

% FIXME: finish migrating the author list
% FIXME: add Gilleain

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collission detection has been
  greatly improved. Figure~\ref{fig:sdg} shows a ...

@@

The chemical formula is the basic/simple chemical representation of a compound.
It defines the number of isotopes or elements that constitute a the composition
of a compound without describing how atoms are bonded. With the rise of metabolomics,
it has become increasingly relevant to have full support in cheminformatics
libraries~\cite{RojasCherto2011}.
It is also the first step
toward the identification of the chemical structure from a unidentified compound.

CDK can handle several concepts related to chemical formulas
\begin{itemize}
\item Chemical formula =
\item Set of chemical formulas =
\item Range of chemical formulas =
\item Adducts =
\item Isotope Container =
\item Isotopic pattern = 
\item Rules = Filters in accepting an chemical formula
\end{itemize}

CDK can
\begin{itemize}
\item generate/print the elemental composition from a given IAtomContainer
\item calculate the isotopic pattern from a given chemical formula
\item determine the elemental composition from a given mass
\item validate a chemical formula
\item calculate the exact mass from a given chemical formula
\end{itemize}

Where can it be applied?

@@

    \begin{minipage}{1\textwidth}
    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
    \centering
    \begin{tabular}{lp{3cm}p{3cm}l}
  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Depedencies} \\ \hline
  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
  standard                   &              & & \\ \hline
  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
  isomorphism                &              & & \\ \hline
  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
  valencycheck               &              & & \\ \hline
  ioformats                  &              & & \\ \hline
  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
  iordf                      &              & New. & Jena 2.7.4 \\ \hline
  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{beam} \\
  smarts                     &              & & Beam 0.9.1~\cite{beam} \\
  formula                    &              & New. & \\ \hline
  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
  qsarmolecular              &              & & \\ \hline
  log4j                      &              & & Log4j 1.2.17 \\ \hline
    \end{tabular}
    \end{minipage}

>>>>

% FIXME: finish migrating the author list
% FIXME: add Gilleain
% FIXME: invite Tomas Pluskal

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collission detection has been
  greatly improved. Figure~\ref{fig:sdg} shows a difference in output between
  the old code base, with and without overlap resolving, and with the new code.
  The new code can be used like:

@@

The chemical formula is the basic/simple chemical representation of a compound.
It defines the number of isotopes or elements that constitute the composition
of a compound without describing how atoms are bonded. With the rise of
metabolomics it has become increasingly relevant to have full support for these
in cheminformatics libraries~\cite{RojasCherto2011}.

The CDK interfaces can handle several concepts related to chemical formulas:
the formula itself, sets of formulas, chemical formula ranges, adducts,
isotope containers and patterns, and rules to filter formula sets. These new
tools can be used for a number of tasks, including calculate the isotopic
pattern from a given chemical formula, determining the possible elemental
compositions for a given mass, and calculate the exact mass from a given
chemical formula.

The following code calculates all possible chemical formula for a given
accurate mass, within some allowed counts for each element:

\begin{verbatim}
Isotopes ifac = Isotopes.getInstance();
MolecularFormulaRange range =
  new MolecularFormulaRange();
range.addIsotope( ifac.getMajorIsotope("C"), 8, 20);
range.addIsotope( ifac.getMajorIsotope("H"), 0, 20);
range.addIsotope( ifac.getMajorIsotope("O"), 0, 1);
range.addIsotope( ifac.getMajorIsotope("N"), 0, 1);

MolecularFormulaGenerator tool =
  new MolecularFormulaGenerator(
    SilentChemObjectBuilder.getInstance(),
    133.0, 133.1, range
  );
IMolecularFormulaSet mfSet = tool.getAllFormulas();
for (mf in mfSet) {
  println MolecularFormulaManipulator.getString(mf) + " " +
    MolecularFormulaManipulator.getTotalExactMass(mf)
}
\end{verbatim}

This gives the following output:

\begin{verbatim}
C11H 133.007825032
C9H11N 133.089149352
C9H9O 133.065339908
C8H7NO 133.052763844
\end{verbatim}

@@

    \begin{minipage}{1\textwidth}
    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
    \centering
    \begin{tabular}{lp{3cm}p{3cm}l}
  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Depedencies} \\ \hline
  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
  standard                   &              & & \\ \hline
  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
  isomorphism                &              & & \\ \hline
  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
  valencycheck               &              & & \\ \hline
  ioformats                  &              & & \\ \hline
  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
  iordf                      &              & New. & Jena 2.7.4 \\ \hline
  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\
  smarts                     &              & & Beam 0.9.1~\cite{Beam} \\
  formula                    &              & New. & \\ \hline
  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
  qsarmolecular              &              & & \\ \hline
  log4j                      &              & & Log4j 1.2.17 \\ \hline
    \end{tabular}
    \end{minipage}
